\documentclass{article} % change to whatever later
\usepackage{fullpage}
\usepackage{amsmath,amsthm}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}

\theoremstyle{definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\PL}{PL} % expand out later
\newcommand{\BSp}{\(\mathbf{B}\)-Species}
\newcommand{\LSp}{\(\mathbf{L}\)-Species}

\title{Structured Positions}
\author{}

\begin{document}
\maketitle
\begin{abstract}
The well-supported data-structures, aka inductive types, in functional
languages are insufficiently rich. We can do better.
\end{abstract}

The starting-point question is: how can one properly teach a \PL about
bags (aka multisets)?

Adding a single new type is rather boring though. Even though doing it
properly is difficult (most languages have rather poor abstraction facilities
so that giving a proper API that doesn't \emph{leak} is hard), it is still
to \textit{ad hoc}. The desire is to find a family of ``type constructors''
that would contain both inductive types and bags.

And even that is aiming far too low. The real problem is 
\textbf{how to teach computers about structures}. In other words, if a
human performs the act of defining a particular structure, a lot of
natural \emph{kit} should come along ``for free'' as a side-effect of
that declaration of intent. (Haskell's \texttt{Deriving} mechanism is
a reasonable analogy; of course getting an induction principle should also
be within scope).

\section{Notation and Vocabulary}

\begin{itemize}
\item we call \emph{container description} what others call \emph{container},
\item we call \emph{container} what others call \emph{structure} or
instance.
\end{itemize}
\noindent Reasoning: we want containers to contain things.

\section{Better Structures}

In a way, we already know that this is \textit{in theory} feasible:
species, \BSp and \LSp. In practice, the theory is insufficiently
constructive nor does it provide answers for key aspects of treating
a structure so obtained as a \emph{data} structure. In particular, no
eliminator.

Another approach is to use the theory of containers. Which, unfortunately,
don't give eliminators either!

\section{Random Jumble}

Basically, self-notes so that things don't get forgotten.

\subsection{Symmetries of trees}
\label{sec:sym-trees}

Start with the not terribly useful "the symmetries of the positions of List".
Boring, as they are trivial. BUT List is tricky as it is the normal form (or
effective presentation) of the carrier of free monoids. You *could* also take
the term algebra and quotient it by the monoid equations (as a classical
mathematician might do). The resulting trees *do* have symmetries! These are
easier to see if one generalizes positions to be 2-sorted, with 1 sort for the
things that contain data, and another for the 'nil' nodes. The resulting
symmetries basically says that the 'nil' nodes cannot convey information. I
don't have an easy way to talk about associativity (yet).  

\subsection{Whence the groups?}

I'm thinking that the group elements (which we now know are equivalence
witnesses for our proof-relevant Setoids of positions) will have a double life,
and show up as generators and relations of an equational theory. Because
equivalences have reflectivity, symmetry and transitivity (and these are
coherent), this is why we get groups. But which group is guided by what other
terms we have. And, lo and behold, all 'swap's is what generates all
permutations is what generates Sn is what gives Bags which are free commutative
monoids.

The connection with Section~\ref{sec:sym-trees} should be clear.

\subsection{Free thingy on X}

Ok, need to code that up the 'silly' way. Induce (single-sorted)
containers from adjunctions between a category of theories and Set.
The counit will inevitably need to respect the equations. The 'containers'
here will be term algebras, modulo the equations. The shapes will be the
terms over Unit; the equations will induce isomorphisms between shapes
as well as between positions.  Unit will be there to mark the holes. The
names of the positions (position labels, what have you) will be the paths
in the structure.

(Something something about counit only taking in 'thingy' homomorphisms
and must only respect those. Need to expand this out.

Really need to code that up, all of the above in this subsection.

\subsection{On eliminators}

Simple desiderata: eliminators should be `compositional' and respect
the appropriate equivalence.

So if a structure naturally decomposes into a disjoint union of
two sub-structures, the eliminator should map. If there is such a thing
as an empty container that belongs to a container description

As Bags decompose as disjoint unions of the 'right' size, its eliminator
must also give the same answer along all such decompositions.

\subsection{Misc}
\begin{itemize}
\item various combinators related to derivatives: up, down, here,
sideways and equinanimity.
\item Ornamentation and differentiation are related: they commute!
\item Ornaments must respect positions (up to iso) but can change shapes
\item Keeping track of where $2$ positions map under the symmetries of
Bag and Cycle will illustrate the difference between them. Only $1$ does not.
\item Specifying by putting a number of fingers down leads to differential
equations on containers.
\item Main trick of dependent types: if you want it respected, put it in
the type.
\item Odd thing about HoTT and UF: they want it respected so much, they put
it into the meta-theory.
\item There seems to be something about the ``directed accessibility
graph'' of the positions of (finitary) containers. It contains both the usual
stuff and, via the non-tree parts, equivalences.
\end{itemize}

\subsection{To remember}
\begin{itemize}
\item Baggy Rose Tree (Philippa Gardner, modelling DOM)
\item Traversal (without leaking) and applicative
\item Bag $\circ$ Maybe $\equiv$ Bag.
\item non-empty container satisfying \texttt{allEqual} $\Rightarrow$ can
extract
\item Species can be seen as generalized vectors with finite sets of
positions.
\item Group actions are way more important than groups alone
\item Want some set of containers closed under differentiation and
preferably some notion of integration too.
\item Cycle(A+B) is a nice structure
\item The proof of the implicit species theorem should be mined fully
\item (Observation from Daan Leijen's talk at POPL): ``Minamide tuples''
are derivatives - and drive Daan's work. Even has a certain FToC feel.
\end{itemize}
\end{document}
